Post-Hartree--Fock methods are often limited by the storage and repeated use of large tensors: wavefunction coefficient objects in configuration-interaction calculations and the electron--repulsion integral (ERI) tensor in perturbative correlation. These tensors are rarely fully dense in an information-theoretic sense, but treating them as strictly sparse by aggressive elementwise truncation can be too crude. Hierarchical compression offers an effective middle ground, combining blockwise sparsity with locally controlled low-rank structure to reduce memory and accelerate dominant contractions while retaining systematic accuracy control.

We develop two hierarchical methods tailored to distinct bottlenecks. For complete active space and full configuration interaction, we introduce corner hierarchically approximated configuration interaction (CHACI), which compresses the CI vector using corner hierarchical partition, norm-based sorting, and blockwise rank selection to maximize information density. By application to dodecacene, a strongly correlated molecule, we demonstrate that CH matrix compression provides superior compression compared to a truncated global singular value decomposition. For Moller-Plesset perturbative methods, we present a hierarchical SOS-MP2 algorithm that applies $\mathcal{H}^{2}$ compression to the ERI tensor within an atomic-orbital Laplace framework, leveraging far-field low rank and sparsity in energy-weighted density matrices. The resulting method achieves $\mathcal{O}(N^{2}\log N)$ time and space complexity for the Coulomb-like SOS-MP2 term and exhibits better-than-$\mathcal{O}(N^{2})$ scaling on linear alkanes and three-dimensional water clusters with $N$ as the number of basis sets.
