First and foremost, I would like to express my deepest gratitude to my advisor, Professor Xiangmin Jiao.
His exceptionally clear vision of the field and patient guidance have been invaluable throughout my Ph.D.\ journey. Under his mentorship, I have significantly strengthened my mathematical foundation and greatly improved my coding skills. His genuine passion for academic research consistently inspired me and helped shape both my research direction and my growth as an independent scholar.

I am also profoundly grateful to my co-advisor, Professor Benjamin Levine, for his insightful guidance in chemistry and for the tremendous effort he devoted to paper writing and responses. I would like to thank Professor Harrison and the Institute for Advanced Computational Science (IACS) for providing funding and, more importantly, for offering opportunities that broadened my academic horizons and enabled meaningful scholarly exchanges. I sincerely thank my thesis chair, Professor Roman Samulyak, and my committee member, Professor Marivi Fernandez-Serra, for their careful reading of my dissertation and for the time, energy, and thoughtful suggestions they contributed to improving this work.

I would like to thank all the present and past members of my research group, including Yipeng Li, Qiao Chen, Xuebin Wang, Jacob Jones, and Chao Zhang, for the many helpful discussions, feedback, and support during group meetings.
I also gratefully acknowledge my collaborators on my Ph.D. projects, including Kenneth Berard and Ying You.
Without their contributions, it would have been much more difficult to complete this work so efficiently.

Finally, I would like to thank my parents for their unconditional love and support. I am especially grateful to my girlfriend, Tianyi Chen, for her understanding, encouragement, and companionship. I also thank my friends in Stony Brook, including Fengchun, Zelong, Yuqi, Zhixing, Chengpeng, Yushen, Wenhan, Sijia, Josh, Zhengxiang, Cheng and many others, for making my life during this time meaningful and joyful.